Synthesizing Quotient Lenses

### 2018-03-13 papers/QuotientLenses/main.tex
finished pass over section 4

\begin{figure}[t]
\centering
\[
\begin{array}{l@{}l@{\ \ }l}
w \; \equiv_{\normalize{R_1}{R_2}{f}} \; w' &\iff&
f(w)=f(w') \\
w \; \equiv_{\id(R)} \; w' &\iff& w = w' \\
w \; \equiv_{\project{R}{s}} \; w' && \text{for all }w, w'\\
w \; \equiv_{\squash{Q_1}{Q_2}{f}} \; w' &\iff& f(w) \equiv_{Q_2} w'
\text{ or } w \equiv_{Q_2} w' \\
w \; \equiv_{Q_1 \; ; \; Q_2} \; w' &\iff& \exists k, k' \in
\mathcal{L}(K(Q_2)) \text{ such that } w \; \equiv_{Q_1} \; k, \; w' \;
\equiv_{Q_1} \; k', \text{ and } k \; \equiv_{Q_2} \; k'\\
w \; \equiv_{Q_1 \cdot Q_2} \; w'  &\iff& w = r_1
\cdot r_2, \; w' = {r'}_1 \cdot {r'}_2 \text{ with } r_1 \; \equiv_{Q_1}
\; {r'}_1, \; r_2 \; \equiv{Q_n} \; {r'}_2\\
w \; \equiv_{Q_1 \sep Q_2} \; w' &\iff& w \; \equiv_{Q_1} \; w'
\text{ or } \; w \; \equiv_{Q_2} \; w'\\
w \; \equiv_{Q^*} \; w' &\iff& w = r_1 \cdot \ldots \cdot r_n, \; w'
= {r'}_1 \cdot \ldots \cdot {r'}_n \text{ and } r_i \equiv_{Q} \; {r'}_i
\\
w \; \equiv_{\perm{Q_1, \ldots, Q_n}{Q}} \; w' &\iff& w = r_{\sigma(1)}
\cdot s_1 \cdot \ldots \cdot s_{n-1} \cdot r_{\sigma(n)}, \;
w' = {r'}_{\theta(1)} \cdot s'_1 \cdot \ldots \cdot s'_{n-1}
\cdot {r'}_{\theta(n)} \\
& & \text{for some } \sigma, \theta \in S_n, \text{ with } r_i \;
\equiv_{q_i} \; r'_i \text{ and } s_k \; \equiv_{Q} \; s'_{k}
\end{array}
\]
\caption{QRE Equivalence Relations}
\label{fig:relations}
\end{figure}
\subsection{Ambiguity and Well Formed QREs}
\label{subsec:well-formed-qres}
An issue that we address before presenting the full set of inference
rules for QREs is the issue of {\em ambiguity}, which appears in the
premises of the inference rules for the regular combinators. If $Q_1$ and $Q_2$
are QREs, then when applying the regular combinators to QREs, we require that
if a string $s$ matches any of the regular expressions $W(Q_1) \cdot W(Q_2)$,
\quad $W(Q_1) \sep W(Q_2)$, \quad $W(Q_1)^*$, \quad $K(Q_1) \cdot K(Q_2)$,
\quad $K(Q_1) \sep K(Q_2)$, \quad $K(Q_1)^*$, then $s$ matches that regular
expression in only one way.

Formally, we say that regular expressions $R$ and $S$ are
\textit{unambiguosly concatenable}, written $R \cdot^! S$ if for all strings
$r, r' \in \mathcal{L}(R)$ and $s, s' \in \mathcal{L}(S)$, if $r \cdot s = r'
\cdot s'$, then $r = r'$ and $s = s'$. We say that a regular expression $R$ is
\textit{unambiguosly iterable}, written $R^{*!}$ if for all strings $r_1,
\ldots, r_m$ and $r'_1, \ldots, r'_n \in \mathcal{L}(R)$, if $r_1 \cdot \ldots
\cdot r_m = r'_1 \cdot \ldots \cdot r'_n$, then $m = n$ and $r_i = r'_i$. We say
that a regular expression $R$ is \textit{strongly unambiguous}\cite{Sippu1988}
if and only if (1) $R = \varnothing$, or (2) $R = S_1 \cdot S_2$ with $S_1,
S_2$ strongly unambiguous and $S_1 \cdot^! S_n$, or (3) $R = S_1 \sep S_2$ with
$S_1, S_2$ strongly unambiguous and $\mathcal{L}(S_1) \cap \mathcal{L}(S_2) =
\varnothing$, or (4) $R = S^*$ with $S$ strongly unambiguous and $S^{*!}$.

Strong unambiguity is necessary in the premises of the regular QRE combinators
because it ensures that the canonizing function of a QRE is well-defined.
For example consider the QRE \lstinline{"a"* . (project "a"* $\to$ "a")}. The
behaviour of this QRE is not-well defined since the string
\cd{"aaa"} can be canonized to any of \cd{"a"}, \cd{"aa"}, or \cd{"aaa"}
depending on how \cd{"aaa"} is parsed. We also require that the regular
combinators applied to the kernels of QREs be unambiguous since the underlying
bijective string lens of a QRE lens operates on kernels, and bijective string lenses
impose the same unambiguity restrictions so that they too are well defined as
functions.

@@

\begin{prooftree}
\AxiomC{$\wf{Q_i, Q}$}
\RightLabel{(Perm)}
\AxiomC{${\substack{\forall \sigma \neq \theta, \; W(c_{\sigma(1)}) \cdot W(Q)
\cdot \ldots \cdot W(c_{\sigma(n)})\\ \cap W(c_{\theta(1)}) \cdot W(Q)
\cdot \ldots \cdot W(c_{\theta(n)}) =\varnothing}}$}
\AxiomC{$K(Q_1) \cdot^! K(Q) \cdot^! \ldots \cdot^! K(Q) \cdot^! K(Q_n)$}
\TrinaryInfC{$\wf{\perm{Q_1, \ldots, Q_n}{Q}}$}
\end{prooftree}

@@

\begin{center}
\AxiomC{$\wf{Q_1, Q_2}$}
\AxiomC{$W(Q_1) \cap W(Q_2) = \varnothing$}
\RightLabel{(Union)}
\BinaryInfC{$\wf{Q_1 \sep Q_2}$}
\DisplayProof
\
\end{center}
\end{center}
\caption{Well-formed QREs}
\label{fig:qrerules}
\end{figure}
As we already mentioned, the unambiguity conditions come into play when defining
QREs using the regular combinators. For example, the (Concat) inference rule
for concatenation says that the concatenation $Q_1 \cdot Q_2$ of QREs $Q_1$ and
$Q_2$ is well formed only if the concatentions of $W(Q_1)$ and $W(Q_2)$, and
$K(Q_1)$ and $K(Q_2)$ are unambiguous.

The most complicated inference rule is the (Perm) rule for the
\kw{permutation} combinator. The second hypothesis for the
Perm rule says that for any two different permutations $\sigma$
and $\theta$, the languages $W(c_{\sigma(1)}) \cdot W(Q) \cdot \ldots \cdot
W(c_{\sigma(n)})$ and $W(c_{\theta(1)}) \cdot W(Q) \cdot \ldots \cdot
W(c_{\theta(n)})$ must be disjoint. This is because any string that matches the
\kw{permutation} combinator matches the language $W(c_{\theta(1)}) \cdot
W(Q) \cdot \ldots \cdot W(c_{\theta(n)})$ for some permutation $\theta$, so we
require that these regular expressions be disjoint.
%
The third hypothesis says that the regular expression $K(Q_1) \cdot K(Q)
\cdot \ldots \cdot K(Q) \cdot K(Q_n)$ must be unambiguous. This is because the
underlying lens that maps to or from a \kw{permutation} QRE operates on
the language $K(Q_1) \cdot K(Q) \cdot \ldots \cdot K(Q) \cdot K(Q_n)$ which is
the kernel of the \kw{permutation} QRE. As we mentioned, the underlying
lens will require that the source and target regular expressions of the lens be
unambiguous so that the lens can match strings uniquely.

>>>>

\begin{figure}[t]
\centering
\[
\begin{array}{l@{}l@{\ \ }l}
w \; \equiv_{\normalize{R_1}{R_2}{f}} \; w' &\iff&
f(w)=f(w') \\
w \; \equiv_{\id(R)} \; w' &\iff& w = w' \\
w \; \equiv_{\project{R}{s}} \; w' &\iff& True \\ % \text{for all }w, w'\\
w \; \equiv_{\squash{Q_1}{Q_2}{f}} \; w' &\iff& f(w) \equiv_{Q_2} w'
\text{ or } w \equiv_{Q_2} w' \\
w \; \equiv_{Q_1 \; ; \; Q_2} \; w' &\iff& \exists k, k' \in
\mathcal{L}(K(Q_2)) \text{ such that } w \; \equiv_{Q_1} \; k, \; w' \;
\equiv_{Q_1} \; k', \text{ and } k \; \equiv_{Q_2} \; k'\\
w \; \equiv_{Q_1 \cdot Q_2} \; w'  &\iff& w = r_1
\cdot r_2, \; w' = {r'}_1 \cdot {r'}_2 \text{ with } r_1 \; \equiv_{Q_1}
\; {r'}_1, \; r_2 \; \equiv{Q_n} \; {r'}_2\\
w \; \equiv_{Q_1 \sep Q_2} \; w' &\iff& w \; \equiv_{Q_1} \; w'
\text{ or } \; w \; \equiv_{Q_2} \; w'\\
w \; \equiv_{Q^*} \; w' &\iff& w = r_1 \cdot \ldots \cdot r_n, \; w'
= {r'}_1 \cdot \ldots \cdot {r'}_n \text{ and } r_i \equiv_{Q} \; {r'}_i
\\
w \; \equiv_{\perm{Q_1, \ldots, Q_n}{Q}} \; w' &\iff& w = r_{\sigma(1)}
\cdot s_1 \cdot \ldots \cdot s_{n-1} \cdot r_{\sigma(n)}, \;
w' = {r'}_{\theta(1)} \cdot s'_1 \cdot \ldots \cdot s'_{n-1}
\cdot {r'}_{\theta(n)} \\
& & \text{for some } \sigma, \theta \in S_n, \text{ with } r_i \;
\equiv_{q_i} \; r'_i \text{ and } s_k \; \equiv_{Q} \; s'_{k}
\end{array}
\]
\caption{QRE Equivalence Relations}
\label{fig:relations}
\end{figure}
\subsection{Ambiguity and Well-Formed QREs}
\label{subsec:well-formed-qres}

To ensure that regular combinations of QREs are well-formed, we need
to enforce a variety of {\em unambiguity} constraints.  
Specifically, when applying regular combinators to QREs $Q_1$ and $Q_2$, we require that
if a string $s$ matches any of the regular expressions $W(Q_1) \cdot W(Q_2)$,
\quad $W(Q_1) \sep W(Q_2)$, \quad $W(Q_1)^*$, \quad $K(Q_1) \cdot K(Q_2)$,
\quad $K(Q_1) \sep K(Q_2)$, \quad $K(Q_1)^*$, then $s$ matches that regular
expression in only one way.
Formally, we say that regular expressions $R$ and $S$ are
\textit{unambiguosly concatenable}, written $R \cdot^! S$ if for all strings
$r, r' \in \mathcal{L}(R)$ and $s, s' \in \mathcal{L}(S)$, if $r \cdot s = r'
\cdot s'$, then $r = r'$ and $s = s'$. We say that a regular expression $R$ is
\textit{unambiguosly iterable}, written $R^{*!}$ if for all strings $r_1,
\ldots, r_m$ and $r'_1, \ldots, r'_n \in \mathcal{L}(R)$, if $r_1 \cdot \ldots
\cdot r_m = r'_1 \cdot \ldots \cdot r'_n$, then $m = n$ and $r_i = r'_i$. We say
that a regular expression $R$ is \textit{strongly unambiguous}\cite{Sippu1988}
if and only if (1) $R = \varnothing$, or (2) $R = S_1 \cdot S_2$ with $S_1,
S_2$ strongly unambiguous and $S_1 \cdot^! S_n$, or (3) $R = S_1 \sep S_2$ with
$S_1, S_2$ strongly unambiguous and $\mathcal{L}(S_1) \cap \mathcal{L}(S_2) =
\varnothing$, or (4) $R = S^*$ with $S$ strongly unambiguous and $S^{*!}$.

Strong unambiguity is necessary in the premises of the regular QRE combinators
because it ensures that the canonizing function of a QRE is well-defined.
For example, consider the QRE \lstinline{"a"* . (project "a"* $\to$ "a")}. The
behaviour of this QRE is not well-defined since the string
\cd{"aaa"} can be canonized to any of \cd{"a"}, \cd{"aa"}, or \cd{"aaa"}
depending on how \cd{"aaa"} is parsed. We also require that the regular
combinators applied to the kernels of QREs be unambiguous since the underlying
bijective string lens of a QRE lens operates on kernels, and bijective string lenses
impose the same unambiguity restrictions so that they too are well defined as
functions.

\subsection{Well-formedness of QREs}

@@

\begin{prooftree}
\AxiomC{$\wf{Q_i, Q}$}
\RightLabel{(Perm)}
\AxiomC{${\substack{\forall \sigma \neq \theta, \; W(Q_{\sigma(1)}) \cdot W(Q)
\cdot \ldots \cdot W(Q_{\sigma(n)})\\ \cap W(Q_{\theta(1)}) \cdot W(Q)
\cdot \ldots \cdot W(Q_{\theta(n)}) =\varnothing}}$}
\AxiomC{$K(Q_1) \cdot^! K(Q) \cdot^! \ldots \cdot^! K(Q) \cdot^! K(Q_n)$}
\TrinaryInfC{$\wf{\perm{Q_1, \ldots, Q_n}{Q}}$}
\end{prooftree}

@@

\begin{center}
\AxiomC{$\wf{Q_1, Q_2}$}
\AxiomC{$W(Q_1) \cap W(Q_2) = \varnothing$}
\RightLabel{(Union)}
\BinaryInfC{$\wf{Q_1 \sep Q_2}$}
\DisplayProof
\
\end{center}
\end{center}
\caption{Well-formed QREs}
\label{fig:qrerules}
\end{figure}
The unambiguity conditions are pertinent when defining QREs using the
regular combinators. For example, the (Concat) inference rule 
says that the concatenation $Q_1 \cdot Q_2$ of QREs
$Q_1$ and $Q_2$ is well formed only if the concatentions of $W(Q_1)$
and $W(Q_2)$, and $K(Q_1)$ and $K(Q_2)$ are unambiguous.

The most complicated inference rule is the (Perm) rule for the
\kw{permutation} combinator. The second hypothesis for the
Perm rule says that for any two different permutations $\sigma$
and $\theta$, the languages $W(Q_{\sigma(1)}) \cdot W(Q) \cdot \ldots \cdot
W(Q_{\sigma(n)})$ and $W(Q_{\theta(1)}) \cdot W(Q) \cdot \ldots \cdot
W(Q_{\theta(n)})$ must be disjoint. This restriction is important
because any string that matches the 
\kw{permutation} combinator matches the language $W(Q_{\theta(1)}) \cdot
W(Q) \cdot \ldots \cdot W(Q_{\theta(n)})$ for some permutation $\theta$, so we
must require that these regular expressions be disjoint.
%
The third hypothesis says that the regular expression $K(Q_1) \cdot K(Q)
\cdot \ldots \cdot K(Q) \cdot K(Q_n)$ must be unambiguous. This
restriction arises because the
underlying lens that maps to or from a \kw{permutation} QRE operates on
the language $K(Q_1) \cdot K(Q) \cdot \ldots \cdot K(Q) \cdot K(Q_n)$, 
the kernel of the \kw{permutation} QRE. The underlying
lens requires that the source and target regular expressions be
unambiguous so that the lens can match strings uniquely.
